\LaTeXe\ guide for the Journal of Functional Programming

### 2019-10-20 papers/jfp/8-engineering.tex
Final tweaks

Another approach is to include failures into the computation context \hs{f}.
Recall from~\S\ref{sec-why-polymorphism} that we require tasks to be polymorphic
in~\hs{f}, and we can therefore choose to execute a \hs{Task} in an~\hs{f} with
failures, the simplest example being \hs{f}~\hs{=}~\hs{Maybe}. Below we define
a callback \hs{fetchA1A2} that returns \hs{Just}~\hs{value} for keys \cmd{A1}
and \cmd{A2} but fails with \hs{Nothing} on all other keys:

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
class Monad m => MonadFail m where
    fail :: String -> m a

sprsh4 :: Tasks MonadFail String Integer
sprsh4 "B1" = Just $ Task $ \fetch -> do
    a1 <- fetch "A1"
    a2 <- fetch "A2"
    if a2 == 0 then fail "division by 0" else return (a1 `div` a2)
sprsh4 _ = Nothing
\end{minted}
\vspace{1mm}

@@

\item The \hs{restarting} scheduler can be made parallel in a few ways, but the
most direct is to have $n$ threads reading keys from the build queue. As before,
if a key requires a dependency not yet built, it is moved to the end~--~the
difference is that sometimes keys will be moved to the back of the queue not
because they are out of date but because of races with earlier tasks that had
not yet finished. As a consequence, if a the calculation order is persisted over
successive runs (as \Excel does), potentially racey dependencies will be
separated, giving better parallelism over time.

@@

Once sufficient parallelism is available the next challenge is preventing excess
parallelism and machine resource starvation, which is usually achieved with a
thread pool or thread limit.

The actual implementation of the parallel schedulers is not overly onerous,
but neither is it beautiful or informative.

@@

Interestingly, there is significant interplay between all three sources of
impurity. Systems like \Bazel use various sandboxing techniques to guard against
missing dependencies, but none are likely to capture all dependencies right down
to the CPU model and microcode version. Tasks that do have untracked
dependencies can be marked as volatile, a technique \Excel takes with the
\cmd{INDIRECT} function, fixing the untracked dependency at the cost of
minimality.

@@

\noindent
Such tasks can be modelled in our framework by adjusting the correctness
definition~(\S\ref{sec-build-correctness}): instead of requiring that the
resulting value \emph{equals the result} of recomputing the task, we now require
that resulting value \emph{belongs to the set of possible results} of
recomputing the task, i.e. the set \cmd{$\{$A1 + 1, A1 + 2$\}$} in the above
example.

@@

Note that \hs{Task}~\hs{MonadRandom} is powerful enough to model
dependency-level non-determinism, for example,
\cmd{INDIRECT("A" \& RANDBETWEEN(1,2))}, whereas most build tasks in real-life
build systems only experience a value-level non-determinism. \Excel handles this
example simply by marking the cell volatile~--~an approach that can be readily
adopted by any of our implementations.

\subsection{Cloud Implementation}\label{sec-cloud-aspects}

@@

\item \textbf{Shallow builds}: Building the end result, e.g. an installer
package, often involves many intermediate tasks. The intermediate results may be
large, so some cloud build systems are designed to build end products
\emph{without materialising} intermediate results, producing a so-called
\emph{shallow build} (see an example in~\S\ref{sec-background-bazel}). Some
build systems go even further, integrating with the file system to only
materialise the file when the user accesses it~\cite{gvfs}.
\end{itemize}

Shallow builds have a slightly weaker correctness property than
Definition~\ref{def-correct}. Instead of demanding that \emph{all} keys reachable
from the target match, we only require matches for the target itself and the \emph{input
keys} reachable from the target.

\begin{figure}
\begin{subfigure}[b]{0.31\linewidth}
\centerline{\includegraphics[scale=0.26]{fig/frankenbuild-example-build.pdf}}
% \vspace{-1.5mm}
\caption{Initial build}
\end{subfigure}
\begin{subfigure}[b]{0.31\linewidth}
\centerline{\includegraphics[scale=0.26]{fig/frankenbuild-example-clean.pdf}}
% \vspace{-1.5mm}
\caption{Clean up, evict \cmd{main.prof}}
\end{subfigure}
\begin{subfigure}[b]{0.36\linewidth}
\centerline{\includegraphics[scale=0.26]{fig/frankenbuild-example-rebuild.pdf}}
% \vspace{-1.5mm}
\caption{Build \cmd{main.prof} and \cmd{report.txt}}
\end{subfigure}
% \vspace{-1.5mm}
\caption{A Frankenbuild example: (a)~build a human-readable profiling report for
\cmd{main.exe} from information dump \cmd{main.prof} produced by a profiling
tool, then record deep constructive traces in the cloud, (b)~remove built files
locally and evict \cmd{main.prof} from the cloud storage, (c)~build
\cmd{main.prof} by executing the profiler (profiling is non-deterministic, hence
the new hash value), then build \cmd{report.txt} by downloading it from the
matching deep constructive trace in the cloud, resulting in a Frankenbuild
because \cmd{main.prof} and \cmd{report.txt} are inconsistent. New and evicted
cloud storage entries are highlighted; file hashes are shown in circles.
\label{fig-frankenbuild}}
% \vspace{-4mm}
\end{figure}

As described in \S\ref{sec-using-cloud-shake}, non-determinism  (\S\ref{sec-non-determinism})
is harmful for cloud builds, reducing the number of cache hits.
However, for deep constructive traces (\S\ref{sec-deep-constructive-traces}) it
is much worse, even leading to \emph{incorrect results}.
Fig.~\ref{fig-frankenbuild} shows a
\emph{Frankenbuild}~\cite{esfahani2016cloudbuild} example, where the target
\cmd{report.txt}, which is downloaded from the cloud, is inconsistent with its
immediate dependency \cmd{main.prof}. This inconsistency is caused by two factors:
(i)~inherent non-determinism of profiling: running a profiling tool on the very
same \cmd{main.exe} will produce different \cmd{main.prof} results every time;
and (ii)~relying on deep constructive traces, which cache build results based
only on the hashes of terminal task inputs (in this case \cmd{main.exe}). Note
that the resulting store is incorrect according to all three definitions of
correctness: the main definition~(\S\ref{sec-build-correctness}), the variant
for non-deterministic tasks~(\S\ref{sec-non-determinism}) and the variant for
shallow builds~(this section).

@@

\subsection{Iterative Computations}\label{sec-iterative-compute}

@@

\subsection{Key-dependent Value Types}\label{sec-polymorphism}

Key-dependent value types allows a single build system to work with multiple different
types of values, where the type of any particular value is determined by the key. As an example
of why this might be useful, consider a build system that can depend on files
(whose contents are strings) or system executables (represented by their version number) --
using a single type for both values reduces type safety.
\Shake permits such key-dependent value types, for example the \emph{oracle} rule from
Mitchell~\shortcite{mitchell2012shake}, and users have remarked that this additional type safety
provides a much easier expression of concepts, e.g. see~Mokhov~\etal~\shortcite{hadrian}.

@@

Some build tasks produce multiple output keys -- for example \cmd{ghc A.hs} produces both
\cmd{A.hi} and \cmd{A.o}. This pattern can be encoded by having a key \cmd{A.hi+A.o} which produces
both results, then each of \cmd{A.hi} and \cmd{A.o} can extract the relevant result
from \cmd{A.hi+A.o}. We can express this pattern more clearly with the extra type safety
from~\S\ref{sec-polymorphism}, using either pairs or lists for both the key and value:

@@

The only awkward aspect is that both \cmd{A.hi} and \cmd{A.o} must ask for
exactly the same \hs{Files} key. If one \hs{File} key swapped the order of the
list to \hs{Files} then it would likely run GHC twice. To help construct a
well-formed multiple-output build task it is convenient to partition the set of
keys into \emph{elementary subsets}. We can encode such a partition with a
function from any key to all the members of its subset (in a consistent order).

@@

In the above function we compute the partition, find the index of this key within that
partition (using \hs{elemIndex}), run the \hs{fetch} on every key in the partition,
then extract out our specific result (using the indexing operation \hs{!!}).

@@

\noindent
The build systems that have precise self-tracking are all ones which
use a \emph{non-embedded domain specific language} to describe build
tasks; that is, a task is described by a data structure (or syntax
tree), and tasks can be compared for equality by comparing those data
structures. Those which use a full programming language, e.g. \Shake,
are faced with the challenge of implementing equality on arbitrary
tasks -- and a task is just a function.  For such build systems, the
only safe approach is to assume (pessimistically) that any
change to the build system potentially changes any build task ~--~ the
classic example being a makefile depending on itself.

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
selfTracking :: (s -> Task Monad k v) -> Tasks Monad k s
                                      -> TasksT Monad (Key k v s)
selfTracking _     _     (Script _) = Nothing -- Scripts are inputs
selfTracking parse tasks (Value  k) = runScript <$> tasks k
  where
    -- Fetch the script, parse it, and then run the obtained task
    runScript :: Task Monad k s -> TaskT Monad (Key k v s) v
    runScript task = TaskT $ \fetch -> do
        script <- run task (fetch . Script)
        run (parse script) (fetch . Value)
\end{minted}
\vspace{1mm}

@@

Most build systems use files for their inputs. When running a build, to check if
a file is up-to-date, one option is to take the modification time or compute the
content hash of each input file. However, for the largest projects, the mere act
of checking the modification time for all the input files can become
prohibitive. To overcome that problem, build systems targeted at large scale,
e.g. \Bazel and \Buck, make use of file watching API's that detect when a file
has changed. Using that information the build can avoid performing many file
accesses.

>>>>

Another approach is to include failures into the computation context \hs{f}.
Recall from~\S\ref{sec-why-polymorphism} that we require tasks to be polymorphic
in~\hs{f}, and can therefore choose to execute a \hs{Task} in an~\hs{f} with
failures, the simplest example being \hs{f}~\hs{=}~\hs{Maybe}. Below we define
a callback that returns \hs{Just}~\hs{value} for keys \cmd{A1} and \cmd{A2} but
fails with \hs{Nothing} on all other keys:

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
class Monad m => MonadFail m where
    fail :: String -> m a
\end{minted}
\begin{minted}[xleftmargin=10pt]{haskell}
sprsh4 :: Tasks MonadFail String Integer
sprsh4 "B1" = Just $ Task $ \fetch -> do
    a1 <- fetch "A1"
    a2 <- fetch "A2"
    if a2 == 0 then fail "division by 0" else return (a1 `div` a2)
sprsh4 _ = Nothing
\end{minted}
\vspace{1mm}

@@

\item The \hs{restarting} scheduler can be made parallel in a few ways, but the
most direct is to have $n$ threads reading keys from the build queue. As before,
if a key requires a dependency not yet built, it is moved to the end~--~the
difference is that sometimes keys will be moved to the back of the queue not
because they are out of date but because of races with earlier tasks that had
not yet finished. As a consequence, if the build order is persisted over
successive runs (as in \Excel), potentially racey dependencies will be
separated, giving better parallelism over time.

@@

Once sufficient parallelism is available the next challenge is preventing excess
parallelism and machine resource starvation, which is usually achieved with a
thread pool/limit.

The actual implementation of the parallel schedulers is not overly onerous, but
neither is it beautiful or informative, so we omit it.

@@

There is significant interplay between all three sources of
impurity. Systems like \Bazel use various sandboxing techniques to guard against
missing dependencies, but none are likely to capture all dependencies right down
to the CPU model and microcode version. Tasks with untracked dependencies can be
marked volatile, a technique \Excel takes with the \cmd{INDIRECT} function,
fixing the untracked dependency at the cost of minimality.

@@

\noindent
Such tasks can be modelled in our framework by adjusting the correctness
definition~(\S\ref{sec-build-correctness}): instead of requiring that the
produced value \emph{equals the result} of recomputing the task, we now require
that produced value \emph{belongs to the set of possible results} of
recomputing the task, i.e. the set \cmd{$\{$A1 + 1, A1 + 2$\}$} in the above
example.

@@

Interestingly, \hs{Task}~\hs{MonadRandom} is powerful enough to express
dependency-level non-determinism, for example,
\cmd{INDIRECT("A" \& RANDBETWEEN(1,2))}, whereas most build tasks in real-life
build systems only experience a value-level non-determinism. \mbox{\Excel}
handles this example simply by marking the cell volatile~--~an approach that can
be readily adopted by any of our implementations.

\vspace{-1mm}
\subsection{Cloud Implementation}\label{sec-cloud-aspects}

@@

\item \textbf{Shallow builds}: Building the end target, e.g. an installer
package, often involves many intermediate tasks. The values produced by these
intermediate tasks may be large, so some cloud build systems are designed to
build end targets \emph{without materialising} any intermediate values,
producing a so-called \emph{shallow build} -- see an example
in~\S\ref{sec-background-bazel}. Some build systems go even further, integrating
with the file system to only materialise the file when the user accesses
it~\cite{gvfs}.
\end{itemize}

Shallow builds have a slightly weaker correctness property than in the
Definition~\ref{def-correct}. Instead of demanding that \emph{all} keys reachable
from the target match, we only require matches for the target itself and the
\emph{input keys} reachable from the target.

\begin{figure}
\begin{subfigure}[b]{0.31\linewidth}
\centerline{\includegraphics[scale=0.27]{fig/frankenbuild-example-build.pdf}}
% \vspace{-1.5mm}
\caption{Initial build}
\end{subfigure}
\begin{subfigure}[b]{0.31\linewidth}
\centerline{\includegraphics[scale=0.27]{fig/frankenbuild-example-clean.pdf}}
% \vspace{-1.5mm}
\caption{Clean up, evict \cmd{main.prof}}
\end{subfigure}
\begin{subfigure}[b]{0.36\linewidth}
\centerline{\includegraphics[scale=0.27]{fig/frankenbuild-example-rebuild.pdf}}
% \vspace{-1.5mm}
\caption{Build \cmd{main.prof} and \cmd{report.txt}}
\end{subfigure}
\vspace{-1mm}
\caption{Frankenbuild example: (a)~build a human-readable profiling report for
\cmd{main.exe} from information dump \cmd{main.prof} produced by a profiling
tool, saving deep constructive traces in the cloud, (b)~remove built files
locally and evict \cmd{main.prof} from the cloud storage, (c)~rebuild
\cmd{main.prof} (profiling is non-deterministic, hence a new hash value), then
build \cmd{report.txt} by downloading it from the matching deep constructive
trace in the cloud. The result is a Frankenbuild because \cmd{main.prof} and
\cmd{report.txt} are inconsistent. New and evicted cloud storage entries are
highlighted; file hashes are shown in circles.
\label{fig-frankenbuild}}
% \vspace{-4mm}
\end{figure}

As described in \S\ref{sec-using-cloud-shake}, non-determinism
(\S\ref{sec-non-determinism}) is harmful for cloud builds, reducing the number
of cache hits. However, for deep constructive traces
(\S\ref{sec-deep-constructive-traces}) it is much worse, even leading to
\emph{incorrect results}. Fig.~\ref{fig-frankenbuild} shows a
\emph{Frankenbuild}~\cite{esfahani2016cloudbuild} example, where the target
\cmd{report.txt}, which is downloaded from the cloud, is inconsistent with its
immediate dependency \cmd{main.prof}. This inconsistency is caused by two factors:
(i)~inherent non-determinism of profiling -- running a profiling tool on the very
same \cmd{main.exe} will produce different \cmd{main.prof} results every time;
and (ii)~relying on deep constructive traces that cache build results based
on the hashes of terminal task inputs (in this case \cmd{main.exe}). The result
violates all three definitions of correctness: the main
definition~(\S\ref{sec-build-correctness}), the variant for non-deterministic
tasks~(\S\ref{sec-non-determinism}) and the variant for shallow
builds~(this section).

@@

\vspace{-1mm}
\subsection{Iterative Computations}\label{sec-iterative-compute}

@@

\vspace{-1mm}
\subsection{Key-dependent Value Types}\label{sec-polymorphism}

Key-dependent value types allow a build system to work with multiple different
types of values, where the type of any particular value is determined by the
key. As an example of why this might be useful, consider a build system where
keys can be files (whose contents are strings) or system executables
(represented by their version number) -- using a single type for both values
reduces type safety. \Shake permits such key-dependent value types, e.g. see
the \emph{oracle} rule in Mitchell~\shortcite{mitchell2012shake}, and users
have remarked that this additional type safety provides a much easier expression
of concepts~\cite{hadrian}.

@@

Some build tasks produce multiple output keys -- for example \cmd{ghc A.hs} produces both
\cmd{A.hi} and \cmd{A.o}. This pattern can be encoded by having a key \cmd{A.hi+A.o} which produces
both results, then each of \cmd{A.hi} and \cmd{A.o} can extract the relevant result
from \cmd{A.hi+A.o}. We can express this pattern more clearly with the extra type safety
from~\S\ref{sec-polymorphism}:

@@

One awkward aspect is that both \cmd{A.hi} and \cmd{A.o} must ask for
exactly the same \hs{Files} key. If one \hs{File} key swapped the order of the
list to \hs{Files} then it would likely run GHC twice. To help construct a
well-formed multiple-output build task it is convenient to partition the set of
keys into \emph{elementary subsets}. We can encode such a partition with a
function from any key to all the members of its subset (in a consistent order).

@@

In the above function we compute the elementary subset \hs{ks} of the given key
\hs{k}, find the index of \hs{k} within the subset (using \hs{elemIndex}), run
the \hs{fetch} to build every key in the subset, then extract out the value
corresponding to \hs{k} (using the indexing operation \hs{!!}).

@@

\noindent
The build systems that have precise self-tracking are all ones which
use a \emph{non-embedded domain specific language} to describe build
tasks; that is, a task is described by a data structure (or syntax
tree), and tasks can be compared for equality by comparing those data
structures. Build systems that use a full programming language, e.g. \Shake,
are faced with the challenge of implementing equality on arbitrary
tasks -- and a task is just a function.  For such build systems, the
only safe approach is to assume (pessimistically) that any
change to the build system potentially changes any build task ~--~ the
classic example being build tasks depending on the makefile itself.

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
selfTracking :: (s -> Task Monad k v) -> Tasks Monad k s
                                      -> TasksT Monad (Key k v s)
selfTracking parse tasks key = case key of
    Script k -> getScript <$> tasks k
    Value  k -> runScript <$> tasks k
  where
    -- Build the script and return it
    getScript :: Task Monad k s -> TaskT Monad (Key k v s) s
    getScript task = TaskT $ \fetch -> run task (fetch . Script)
    -- Build the script, parse it, and then run the obtained task
    runScript :: Task Monad k s -> TaskT Monad (Key k v s) v
    runScript task = TaskT $ \fetch -> do
        script <- run task (fetch . Script)
        run (parse script) (fetch . Value)
\end{minted}
\vspace{1mm}

@@

Most build systems use files for their inputs. When running a build, to check if
a file is up-to-date, one option is to take the modification time or compute the
content hash of each input file. However, for the largest projects, the mere act
of checking the modification time for all the input files can become
prohibitive. To overcome that problem, build systems targeted at large scale,
e.g. \Bazel and \Buck, rely on file watching API's to detect when a file has
changed. Using that information the build can avoid performing many file
accesses.

### 2019-10-20 papers/jfp/9-related.tex
Add a TODO
Say more about Redo vs Shake
Final tweaks

\section{Related work}\label{sec-related}

@@

\begin{itemize}
\item \Dune~\cite{dune} is a build system designed for OCaml/Reason projects.
Its distinguishing feature is that it uses
\emph{arrows}~\cite{hughes2000generalising} rather than monads to model
dynamic dependencies, which simplifies static dependency approximation.

@@

\item \Pluto~\cite{erdweg2015pluto} is based on a similar model to \Shake, but
additionally allows cyclic build rules combined with a user-specific resolution
strategy. Often such a strategy can be unfolded into the user rules without loss
of precision, but a fully general resolution handler extends the \hs{Task}
abstraction with additional features.

\item \Redo~\cite{redo-idea,grosskurth2007redo,redo} almost exactly
matches \Shake at the level of detail given here, differing only on aspects like
rules producing multiple files~\S\ref{sec-polymorphism}.

@@

\noindent
To achieve minimality, each separate command is traced at the OS-level, allowing
\Fabricate to record a trace entry stating that \cmd{gcc -c util.c} reads from
\cmd{util.c}. In future runs \Fabricate runs the script from start to finish,
skipping any commands where no inputs have changed. The key difference from our
\hs{Tasks} abstraction is that instead of supplying a mapping from outputs to
tasks, \Fabricate supplies a list of statements, in an order, without declaring
what each line produces. There is no need to schedule the statements, and not
enough information to do so.

Taking our abstraction, it is possible to encode \Fabricate assuming that
commands like \cmd{gcc -c util.c} are keys, there is a linear dependency between
each successive key, and that the OS-level tracing can be lifted back as a
monadic \hs{Task} function\footnote{\Shake provides support for
\Fabricate{}-like build systems~--~see \cmd{Development.Shake.Forward}.}.
However, in our pure model the mapping is not perfect as \cmd{gcc} writes to
arbitrary files whose locations are not known in advance.

@@

While not typically considered build systems, self-adjusting computation is a
well studied area, and in particular the contrast between different formulations
has been thoroughly investigated, e.g.
see~Acar~\etal~\shortcite{acar2007selfadjusting}. Self-adjusting computations
can automatically adjust to an external change to their inputs. A classic
example is a self-adjusting sorting algorithm, which can efficiently (in
$O(\log{n})$ time where $n$ is the length of the input) recalculate the result
given an incremental change of the input. While very close to build systems in
spirit, self-adjusting computations are mostly used for in-memory computation
and rely on the ability to dynamically allocate new keys in the store for
sharing intermediate computations~--~an intriguing feature rarely seen in build
systems (\Shake's oracles~\S\ref{sec-polymorphism} can be used to model this
feature to a limited degree).

A lot of research has been dedicated to finding efficient data structures and
algorithms for self-adjusting computations (with a few open-source
implementations, e.g. \Incremental by Jane Street~\shortcite{incremental}. We
plan to investigate how these insights can be utilised by build systems as
future work.

@@

\emph{Memoization} is a classic optimisation technique for storing values of a
function instead of recomputing them each time the function is called. Minimal
build systems (see the Definition~\ref{def-minimal}) certainly perform
memoization: they \emph{store values instead of recomputing them each time}.
Memoization can therefore be reduced to a minimal build system (as we
demonstrate below), but not vice versa, since minimal build systems solve a more
complex optimisation problem.

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
data Key = A Int | B Int | C Int Int deriving Eq
\end{minted}
\vspace{0mm}
\begin{minted}[xleftmargin=10pt]{haskell}
editDistance :: Tasks Monad Key Int
editDistance (C i 0) = Just $ Task $ const $ pure i
editDistance (C 0 j) = Just $ Task $ const $ pure j
editDistance (C i j) = Just $ Task $ \fetch -> do
    ai <- fetch (A i)
    bj <- fetch (B j)
    if ai == bj
        then fetch (C (i - 1) (j - 1))
        else do insert  <- fetch (C  i      (j - 1))
                delete  <- fetch (C (i - 1)  j     )
                replace <- fetch (C (i - 1) (j - 1))
                return (1 + minimum [insert, delete, replace])
editDistance _ = Nothing
\end{minted}
\vspace{1mm}

\noindent
When asked to build \hs{C}~$n$~$m$, a minimal build system will calculate the
result using memoization. Furthermore, when an input symbol $a_i$ is changed,
only necessary, incremental recomputation will be performed~--~an optimisation
that cannot be achieved just with memoization.

>>>>

\section{Related Work}\label{sec-related}

@@

\begin{itemize}
\item \Dune~\cite{dune} is a build system designed for OCaml/Reason projects.
Its original implementation used \emph{arrows}~\cite{hughes2000generalising}
rather than monads to model dynamic dependencies, which simplified static
dependency approximation. \Dune was later redesigned to use a flavour of
selective functors~\cite{mokhov_selective_2019}, making it a closer fit to our
abstractions.

@@

\item \Pluto~\cite{erdweg2015pluto} is based on a similar model to \Shake, but
additionally allows cyclic build rules combined with a user-specific resolution
strategy. Often such a strategy can be unfolded into the user rules without loss
of precision, but a fully general resolution handler extends the \hs{Task}
abstraction with new features.

\item \Redo~\cite{redo-idea,grosskurth2007redo,redo} almost exactly
matches \Shake at the level of detail given here, differing only on aspects like
rules producing multiple files~\S\ref{sec-polymorphism}. While \Redo predates
\Shake, they were developed independently; we use \Shake as a prototypical
example of a monadic build system because its implementation presents a closer
mapping to our \hs{Task} abstraction.

@@

\noindent
To achieve minimality, each separate command is traced at the OS-level, allowing
\Fabricate to record a trace entry stating that \cmd{gcc -c util.c} reads from
\cmd{util.c}. In future runs \Fabricate runs the script from start to finish,
skipping any commands where no inputs have changed. The main difference from our
\hs{Tasks} abstraction is that instead of supplying a mapping from keys to
tasks, a \Fabricate script supplies a list of build statements, in a
user-scheduled order, without declaring what each statement reads or write.

Taking our abstraction, it is possible to encode \Fabricate assuming that
commands like \cmd{gcc -c util.c} are keys, there is a linear dependency between
each successive key, and that the OS-level tracing can be lifted back as a
monadic \hs{Task} function\footnote{\Shake provides support for
\Fabricate{}-like build systems~--~see \cmd{Development.Shake.Forward}.}.
However, in our pure model the mapping is not perfect as \cmd{gcc} writes to
arbitrary files whose locations are not known in advance. One way of capturing
arbitrary writes in our model is to switch from one callback \hs{fetch} to
\emph{two callbacks}, say \hs{read} and \hs{write}, allowing us to track both
reads and writes separately.

@@

While not typically considered build systems, self-adjusting computation is a
well studied area, and in particular the contrast between different formulations
has been thoroughly investigated, e.g.
see~Acar~\etal~\shortcite{acar2007selfadjusting}. Self-adjusting computations
can automatically adjust to an external change to their inputs. A classic
example is a self-adjusting sorting algorithm, which can efficiently (in
$O(\log{n})$ time where $n$ is the length of the input) recalculate the result
given an incremental change of the input. While very close to build systems in
spirit, self-adjusting computations are mostly used for in-memory computation
and rely on the ability to dynamically allocate new keys in the store for
sharing intermediate computations~--~an intriguing feature rarely seen in build
systems (\Shake's oracles~\S\ref{sec-polymorphism} can be used to model this
feature to a limited degree). Another important optimisation that self-adjusting
computation engines often support is the incremental processing of
\emph{deltas}, where instead of marking a value as ``changed to 8'', one can
mark it as ``changed by $+1$'', assuming it was equal to 7 before. When a delta
is small, it can often be propagated to the output more efficiently than by
recomputing the output value from scratch.

A lot of research has been dedicated to finding efficient data structures and
algorithms for self-adjusting computations, with a few open-source
implementations, e.g. \Incremental by Jane Street~\shortcite{incremental}. We
plan to investigate how these insights can be utilised by build systems as
future work.

@@

\emph{Memoization} is a classic optimisation technique for storing values of a
function instead of recomputing them each time the function is called. Minimal
build systems (\S\ref{sec-background-make}) certainly perform
memoization: they \emph{store values instead of recomputing them each time}.
Memoization can therefore be reduced to a minimal build system (as we
demonstrate below), but not vice versa, since minimal build systems solve a more
complex optimisation problem.

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
data Key = A Int | B Int | C Int Int deriving Eq
\end{minted}
\begin{minted}[xleftmargin=10pt]{haskell}
editDistance :: Tasks Monad Key Int
editDistance (C i 0) = Just $ Task $ const $ pure i
editDistance (C 0 j) = Just $ Task $ const $ pure j
editDistance (C i j) = Just $ Task $ \fetch -> do
    ai <- fetch (A i)
    bj <- fetch (B j)
    if ai == bj
        then fetch (C (i - 1) (j - 1))
        else do insert  <- fetch (C  i      (j - 1))
                delete  <- fetch (C (i - 1)  j     )
                replace <- fetch (C (i - 1) (j - 1))
                return (1 + minimum [insert, delete, replace])
editDistance _ = Nothing
\end{minted}
\vspace{1mm}

\noindent
When asked to build \hs{C}~$n$~$m$, a minimal build system will calculate the
result using memoization. Furthermore, when an input $a_i$ is changed, only
necessary, incremental recomputation will be performed~--~an optimisation that
cannot be achieved just with memoization.
